Topological Equivariant Artist Model

### 2021-03-08 slides/main.tex
shelve intro for now

\documentclass[xcolor={dvipsnames}, handout]{beamer}
\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
\usepackage{mathrsfs} % math script fonts
\usepackage{tikz-cd} % commutative diagrams
\usepackage{subcaption} % subfigure float captioning

@@

\title{Topological Artist Model}

@@

\begin{frame}{What do visualization libraries do?}
    \begin{figure}
        \includegraphics[width=1\linewidth]{figures/intro/dar.png}
    \end{figure}
\end{frame}

\begin{frame}{Why are we doing this?}
    \begin{enumerate}
        \item visualization software design patterns are tuned to data structures\cite{HeerSoftware2006}
        \item visualization types tuned to data structures\cite{toryRethinkingVisualizationHighlevel2004}
        \item need to build a tool that supports domain specific structures and algorithms
        \item tool needs to be general purpose enough to support lots of domains
        \item tools needs to support 2D, 3D, static, dynamic, and interactive viz
    \end{enumerate}
\end{frame}

\subsection{Tools}
\begin{frame}{Everything is a table}
\begin{columns}
    \column{0.5\textwidth}
    \begin{figure}
        \includegraphics[width=1\textwidth]{figures/intro/grammar_example.png}
        \caption{Figure 1.5 in Wilkenson's Grammar of Graphics\cite{wilkinsonGrammarGraphics2005}}
    \end{figure}
    \column{0.5\textwidth}
        \begin{enumerate}
            \item  ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009}, D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, altair\cite{vanderplasAltairInteractiveStatistical2018}
            \item tends to be declarative\cite{heerDeclarative2010} 
            \item users describe how to compose visual elements \cite{wongsuphasawatNavigatingWideWorld2021}
        \end{enumerate}
    \end{columns}
    \pause
    \begin{center}
        \textbf{TAM: build the visual elements}
    \end{center}
\end{frame}

\begin{frame}{Everything is an image}
    \begin{columns}
        \column{0.5\textwidth}
        \begin{figure}
            \includegraphics[width=1\textwidth]{figures/intro/landsat.png}
        \end{figure}
        \column{0.5\textwidth}
        \begin{enumerate}
            \item ImageJ\cite{schneiderNIHImageImageJ2012}, ImagePlot\cite{studiesCulturevisImageplot2021}, Napari\cite{nicholas_sofroniew_2021_4533308}
            \item build plugins into existing system where image is primary input
        \end{enumerate}
    \end{columns}
    \pause
    \begin{center}
        \textbf{TAM: more general than tables and images}
    \end{center}
\end{frame}

\begin{frame}{Everything is ?}
    \begin{columns}
        \column{0.5\textwidth}
        \begin{figure}
            \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
            \caption{Image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
        \end{figure}
        \column{0.5\textwidth}
        \begin{enumerate}
            \item Matplotlib\cite{hunterMatplotlib2DGraphics2007},  VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk}, MayaVi\cite{RamachandranMayaVI2011}, ParaView\cite{ahrens2005paraview}, Titan\cite{brianwylieUnifiedToolkitInformation2009}
            \item each visualization type API is tuned to the data structure
        \end{enumerate}
    \end{columns}
    \pause
    \begin{center}
        \textbf{TAM: common data abstraction interface between data and viz }
    \end{center}
\end{frame}

\subsection{Data}
\begin{frame}{Variables \& Topology}
    \begin{figure}
        \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
        \caption{Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
    \end{figure}
    \begin{description}
        \item  metadata are \textit{keys} with associated \textit{values} (Munzner \cite{munznerVisualizationAnalysisDesign2014})
        \item[topology] Fiber bundles can be a common data abstraction (Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989})
        \item[variables] Fiber as a database schema (Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES})
    \end{description}
    \pause
    \begin{center}
        \textbf{Tam: variables are \textit{values} and \textit{keys} locate them in topology}
    \end{center}
\end{frame}

\subsection{Visualization}
\begin{frame}{Property Matching}
    \begin{figure}
        \includegraphics[width=.7\textwidth]{figures/intro/retinal_variables.png}
        \caption{This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from Krygier and Wood's \textit{Making Maps: A Visual Guide to Map Design for GIS}\cite{krygierMakingMapsVisual2005}}
    \end{figure}
\end{frame}

\begin{frame}{Evaluating visualizations}
    \begin{description}
        \item[Expressiveness] structure preserving mappings from data to graphic (Mackinlay \cite{mackinlayAutomatingDesignGraphical1986})
        \item[Effectiveness] design choices made in deference to perceptual saliency \cite (Mackinlay \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014})
        \item[Naturalness] easier to understand when properties match (Norman \cite{norman_things_smart})
        \item[Graphical Integrity] graphs show \textbf{only} the data (Tufte \cite{tufteVisualDisplayQuantitative2001})
    \end{description}
\end{frame}

\begin{frame}{Mathematical Frameworks for evaluating visualization}
    \begin{enumerate}
        \item APT: visualization has syntax and semantics like a language  (Mackinlay  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987})
        \item Visualization has functional dependencies that can be represented as graphs (Sugibuchi \cite{sugibuchiFramwork2009}) 
        \item the semiotics of visualization are commutative in a category theory framework (Vickers \cite{vickersUnderstandingViz2013})
    \end{enumerate}
    \pause
    \begin{center}
        \textbf{Tam: Framework for building visualizations}
    \end{center}
\end{frame}

\begin{frame}{Visualization is commutative maps}
    data ($\alpha$) and viz ($\omega$) symmetries (Kindlmann and Scheidegger \cite{kindlmann2014algebraic})
        \begin{block}{$v\circ r_2 \circ \alpha =\omega\circ v\circ r_1$}
            \begin{equation*}
            \begin{tikzcd}[ampersand replacement=\&]
                D \arrow[d, "\alpha"'] \arrow[r, "r_1"] \& R \arrow[r, "\nu"]  \& V \arrow[d, "\omega"] \\
                D \arrow[r, "r_2"']                     \& R \arrow[r, "\nu"'] \& V                    
            \end{tikzcd}
        \end{equation*}
    \end{block}
    \pause
    \begin{center}
        \textbf{Tam: Add topology and make it functional}
    \end{center}
\end{frame}
\begin{frame}{So why?}
    \begin{figure}[H]
        \begin{subfigure}{.24\textwidth}
            \includegraphics[width=1\textwidth]{figures/intro/du_bois_spinny.png}
            \caption{}
            \label{fig:intro_dpa}
        \end{subfigure}
        \begin{subfigure}{.24\textwidth}
            \includegraphics[width=1\textwidth]{figures/intro/du_bois_line.png}
            \caption{}
            \label{fig:intro_dpb}
        \end{subfigure}
        \begin{subfigure}{.24\textwidth}
            \includegraphics[width=1\textwidth]{figures/intro/du_bois_country.png}
            \caption{}
            \label{fig:intro_dbc}
        \end{subfigure}
        \begin{subfigure}{.24\textwidth}
            \includegraphics[width=1\textwidth]{figures/intro/du_bois_heat.png}
            \caption{}
            \label{fig:intro_dbd}
        \end{subfigure}
        \caption{Du Bois' data portraits from the Prints and Photographs collection of the Library of Congress \cite{duboisGeorgiaNegroCity1900,duboisGeorgiaNegroValuation1900, duboisSeriesStatisticalCharts, duboisSeriesStatisticalChartsa}}
        \label{fig:intro_dubois}
    \end{figure}
    \begin{center}
        \textbf{Framework for building any valid visualization}
    \end{center}
\end{frame}

\subsection{Contributions}
\begin{frame}{Contributions}
    \begin{enumerate}
        \item a formal description of the topology preserving relationship between data and graphic via continuous maps
        \item a formal description of the property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
        \item abstraction of data structure using fiber bundles with schema like fibers to encode components and topology 
        \item algebraic sum operator such that more complex visualizations can be built from simple ones 
        \item a functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
        \item a prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
    \end{enumerate}
\end{frame}

@@

\begin{frame}{Topological Artist Model}
    \begin{equation}
        \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
    \end{equation}
    \pause
    which is a map from data \dtotal\ to graphic \gtotal\ that carries a homomorphism of monoid actions 
    \pause 
    \begin{equation}
    \varphi: \monoid \rightarrow \monoid^{\prime}
    \end{equation}
    \pause
    such that artists are equivariant maps 
    \begin{equation}
    \mathscr{\vartist}(m\cdot \delement) = \varphi(m)\cdot\mathscr{\vartist}(\delement) 
    \end{equation}
\end{frame}

@@

\begin{frame}{Fiber Components}
    We can decouple decouple \dfiber\ into components $\dfiber_i$
    \begin{equation}
        \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
    \end{equation}
\end{frame}

@@

\begin{frame}{Monoid Composition: Partial Order}
    MAKE THIS Hasse Diagram (borrow from theories of composablity)
\end{frame}

@@

\subsection{Graphics}
\begin{frame}{Fiber Bundle}
    The graphics bundle is a tuple $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$ defined by the projection map $\pi$
    \begin{equation}
        \label{eq:fiber_bundle}
        \begin{tikzcd}[ampersand replacement=\&]
            \gfiber \arrow[r, hook] \& \gtotal \arrow[r, "\pi"] \& \gbase
        \end{tikzcd}
    \end{equation}
\end{frame}

\begin{frame}{Continuity: Base space}
    The surjective map $\vindex: \gbase \rightarrow \dbase$ 
    \begin{equation}
        \begin{tikzcd}[ampersand replacement=\&]
            \dtotal \arrow[d, "\pi"'] \& \gtotal \arrow[d, "\pi"'] \\
            \dbase                   \& \gbase \arrow[l, "\vindex"']
        \end{tikzcd}
    \end{equation}
    goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$. 
\end{frame}

@@

\begin{frame}[allowframebreaks]{References}
\printbibliography
\end{frame}
\end{document}

>>>>

\documentclass[xcolor={dvipsnames}, handout]{beamer}

\usepackage{amsmath,amsfonts,amssymb,pxfonts,eulervm,xspace}
\usepackage{mathrsfs} % math script fonts
\usepackage{tikz-cd} % commutative diagrams
\usepackage{subcaption} % subfigure float captioning

@@

\title{Topological Equivariant Artist Model}

@@

\begin{frame}{Visualizations are structure preserving maps}
    \begin{figure}
        \includegraphics[width=1\linewidth]{figures/intro/dar.png}
    \end{figure}
    The aim of this work is to rearchitecture Matplotlib
    \pause 
    to take advantage of developments in software design, data structures, and visualization
    \pause
    to improve consistency, reusability, and discoverability,
    \pause
    so domain specific tool developers can build structure preserving visualization tools.
\end{frame}

\begin{frame}{Visualization component constraints}
    \begin{description}
        \item[equivariance] properties of data and visual encoding match
        \item[continuity] connectivity of data and visual encoding match
        \item[composibility] structure preserved by individual components is preserved in combined components
    \end{description}
\end{frame}

\subsection{Background}
\begin{frame}{Tools are tuned to the continuity of the date \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}}
    \begin{columns}
        \column{0.33\textwidth}
        \begin{figure}
            \includegraphics[width=1\textwidth]{figures/intro/data_formatting.png}
            \caption{Based on fig~2.5 in Munzner's VAD\cite{VisualizationAnalysisDesign}}
        \end{figure}

        \begin{enumerate}
            \item ggplot\cite{wickhamGgplot2ElegantGraphics2016a}
            \item protovis\cite{bostockProtoviz2009}, D3 \cite{bostockDataDrivenDocuments2011}
            \item vega\cite{satyanarayanDeclarativeInteractionDesign2014}, altair\cite{vanderplasAltairInteractiveStatistical2018}
        \end{enumerate}
        \pause
        \column{0.33\textwidth}
        \begin{figure}
            \includegraphics[width=1\textwidth]{figures/intro/landsat.png}
        \end{figure}
        \begin{enumerate}
            \item ImageJ\cite{schneiderNIHImageImageJ2012}, ImagePlot\cite{studiesCulturevisImageplot2021}\item Napari\cite{nicholas_sofroniew_2021_4533308}
        \end{enumerate}
        \pause
        \column{0.33\textwidth}
        \begin{figure}
            \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
            \caption{Data Representation, MayaVi 4.7.2 docs\cite{DataRepresentationMayavi}}
        \end{figure}
        \begin{enumerate}
            \item Matplotlib\cite{hunterMatplotlib2DGraphics2007}, 
            \item VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk}, MayaVi\cite{RamachandranMayaVI2011}, ParaView\cite{ahrens2005paraview}, Titan\cite{brianwylieUnifiedToolkitInformation2009}
        \end{enumerate}
    \end{columns}
\end{frame}

\begin{frame}{Structure is encoded in variables and continuity}
    \begin{figure}
        \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
        \caption{Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
    \end{figure}
    \begin{description}
        \item[binding] metadata are structural \textit{keys} with associated \textit{values} (Munzner \cite{munznerVisualizationAnalysisDesign2014})] 
        \item[continuity] Fiber bundles can be a common data abstraction (Butler \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989})
        \item[variables] Fibers can hold schema like encodings of variables (Spivak \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES})
    \end{description}
\end{frame}

\begin{frame}{Visualizations are (mostly) evaluated on equivariance}
    \begin{description}
        \item[Expressiveness] structure preserving mappings from data to graphic (Mackinlay \cite{mackinlayAutomatingDesignGraphical1986})
        \item[Effectiveness] design choices made in deference to perceptual saliency \cite (Mackinlay \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014})
        \item[Naturalness] easier to understand when properties match (Norman \cite{norman_things_smart})
        \item[Graphical Integrity] graphs show \textbf{only} the data (Tufte \cite{tufteVisualDisplayQuantitative2001})
    \end{description}
\end{frame}

\begin{frame}{Models describe composition}
    \begin{description}
        \item[language model] APT: syntax and semantics of visualization (Mackinlay  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987})
        \item[functional dependencies] constrained maps between data and visual representation(Sugibuchi \cite{sugibuchiFramwork2009}) 
        \item[category theory] the semiotics of visualization are commutative (Vickers \cite{vickersUnderstandingViz2013})
        \item[algebraic process] data ($\alpha$) and viz ($\omega$) transforms are symmetric (Kindlmann and Scheidegger \cite{kindlmannAlgebraicProcessVisualization2014})
        \begin{columns}
            \column{.5\textwidth}

            \begin{description}
                \item[D] data 
                \item[R] representations
                \item[V] visualizations
            \end{description}
            \column{.5\textwidth}
            \begin{equation*}
                \begin{tikzcd}[ampersand replacement=\&]
                    D \arrow[d, "\alpha"'] \arrow[r, "r_1"] \& R \arrow[r, "\nu"]  \& V \arrow[d, "\omega"] \\
                    D \arrow[r, "r_2"']                     \& R \arrow[r, "\nu"'] \& V                    
                \end{tikzcd}
                \end{equation*}

        \end{columns}

    \end{description}
\end{frame}

\subsection{Contributions}
\begin{frame}{Contributions}
    \begin{description}
        \item[Topological] topology preserving relationship between data and graphic via continuous maps
        \item[Equivariant] property preservation from data component to visual representation as equivariant maps that carry a homomorphism of monoid actions
        \item[Artist] functional oriented visualization tool architecture built on the mathematical model to demonstrate the utility of the model
        \item [Model] prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model
    \end{description}
\end{frame}

@@

\begin{frame}{Topological Equivariant Artist Model}
    The Artist $\mathscr{\vartist}$ is a map from data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$
    \begin{equation}
        \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
    \end{equation}
    \pause
     that carries a homomorphism of monoid actions 
    \begin{equation}
    \varphi: \monoid \rightarrow \monoid^{\prime}
    \end{equation}
    \pause
    such that artists are equivariant maps 
    \begin{equation}
    \mathscr{\vartist}(m\cdot \delement) = \varphi(m)\cdot\mathscr{\vartist}(\delement) 
    \end{equation}
    \pause 
    with a deformation retraction from graphic to data space. 
    \note{some kinda words about the continuity}
\end{frame}

@@

@@

\begin{frame}{Monoid Composition: Partial Order}
    \begin{figure}
        \begin{overprint}
            \onslide<1|handout:0>\includegraphics[width=1\linewidth]{figures/math/monoid_hasse.png}
            \onslide<2|handout:0>\includegraphics[width=1\linewidth]{figures/math/monoid_monotone.png}
            \onslide<3>\includegraphics[width=1\linewidth]{figures/math/monoid_maps.png}
        \end{overprint}
    \caption{Inspired by definition 1.59 diagram in Spivak and Fong's An Invitation to Applied Category Theory \cite{fongInvitationAppliedCategory2019}}
    \end{figure}
\end{frame}

@@

\subsection{Graphics}
\begin{frame}{Graphic Bundle}
    The graphics bundle is a tuple $(\gtotal,\,\gbase,\,\pi ,\,\gfiber)$ defined by the projection map $\pi$
    \begin{equation}
        \label{eq:fiber_bundle}
        \begin{tikzcd}[ampersand replacement=\&]
            \gfiber \arrow[r, hook] \& \gtotal \arrow[r, "\pi"] \& \gbase
        \end{tikzcd}
    \end{equation}
\end{frame}
\begin{frame}{Continuity: Base space}
    The surjective map $\vindex: \gbase \rightarrow \dbase$ 
    \begin{equation}
        \begin{tikzcd}[ampersand replacement=\&]
            \dtotal \arrow[d, "\pi"'] \& \gtotal \arrow[d, "\pi"'] \\
            \dbase                   \& \gbase \arrow[l, "\vindex"']
        \end{tikzcd}
    \end{equation}
    goes from region $\gbasepoint \in \gbase_{\dbasepoint}$ to its associated point $\gbasepoint$. 
\end{frame}

@@

\subsection{Artist}

\section{References}
%\begin{frame}[allowframebreaks]{References}
%\printbibliography
%\end{frame}

\end{document}
